\textsf{jaxoplanet}: Hardware\,-Accelerated Orbits and Light Curves

### 2025-01-14 latex/paper.tex
Add part of dan's comments and So comments

% author info
\author[0000-0002-4296-2246]{Lionel J. Garcia}
\author{Soichiro Hattori}
\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY, USA}

@@

\begin{abstract}
    Modeling light curves of celestial bodies serves a wide variety of science cases. From the orbital characterization of stellar systems to the study of exoplanet and stellar atmospheres. In this context, the analytical framework known as \textit{starry} provided the community with a powerful tool to analytically compute light curves of bodies with non-uniform surface intensities. However, as datasets and models complexity grow, the tractable inference of their physical parameters becomes increasingly challenging, and often motivates biased approximations. In this paper, we present a new implementation of the \textsf{starry} C++ implementation in \textsf{JAX}, a high-performance machine-learning library designed for large-scale computations. We describe the changes made to the original version of \textsf{starry} and evaluate the performance of our new implementation, that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a differentiable model of stellar light curves, and enable the use of powerful machine-learning tools available in the \textsf{JAX} ecosystem. \gitlink{}
\end{abstract}

@@

In the age of wide-field surveys (such as the \textit{Transiting Exoplanet Survey Satellite}; \citealt{tess}) and high-precision spectroscopic instruments, these models are being compared to large datasets in probabilistic ways, requiring fast and robust evaluations of their posterior likelihood over larger and larger parameter spaces. These inferences can be highly facilitated by gradient-based optimization and sampling algorithms (see for example \citealt{Betancourt2017}), so that being able to differentiate these models with respect to their parameters offers clear advantages.\\\\
In this context, \cite{starry} and \cite{Agol2020} presented a framework to analytically compute the light curves of spherical bodies with non-uniform surface intensities, such as stars and exoplanets, and the flux observed from their mutual occultations. This framework, known as \textit{starry}, has been successfully applied to a wide variety of science cases, like the study of transiting exoplanets \citep{Kostov2019}, precise modeling of radial velocities \citep{Dawson2021}, the study of exoplanets atmospheres \citep{Bell2023}, the study of stellar photospheres \citep{Almenara2022}, or even the surface mapping of stellar system objects \citep{Bartolic2022}. However, as datasets and models complexity grow, the tractable inference of the larger parameter spaces we have to deal with becomes increasingly challenging, and often motivates biased approximations. This is the case, for example, in transmission spectroscopy measurements conducted with the \textit{James Webb Space Telescope} NIRSpec instrument, that should ideally consist in the simultaneous modeling of more than 3000 transit light curves, one per pixel element (e.g. \citealt{Alderson2023}).\\\\
In this paper we present \textsf{jaxoplanet}, a new implementation of the \textit{starry} framework in \textsf{JAX}, a high-performance machine-learning library designed for large-scale computations, including autodifferentiation capabilities, with code compiled and deployable on various hardware accelerators. In \autoref{starry}, we summarize the formalism of \textit{starry}, introduced by \cite{starry} to analytically compute stellar light curves. In \autoref{optimization}, we describe the \textsf{jaxoplanet} implementation and the changes made to the \textsf{starry} implementation. Finally, in \autoref{performance}, we evaluate the performance of our new implementation that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a differentiable model of stellar light curves, and enable the use of powerful machine-learning tools available in the \textsf{JAX} ecosystem.\\\\

@@

Inspired by the idea of \cite{pal2012}, \cite{starry} and \cite{Agol2020} present a general framework to compute the light curves of spherical bodies with non-uniform surface intensities. In this section, we describe the mathematical formalism behind the \textit{starry} model.

@@

In order to simplify this integral, \citealt{starry} introduce a change of basis from the spherical harmonics basis $\bvec{\tilde y}$ to a simple polynomial basis $\bvec{\tilde p}$ (defined in \citealt[section 2.3]{starry}) characterized by the change of basis matrix $\bvec{A_1}$. In this new basis, $\bvec{y}$ can be expressed as
\begin{equation}\label{eq:p}\bvec{p} = \bvec{A_1}\bvec{y},\end{equation}
such that 

@@

As $\bvec{\tilde{p}}$ is simply made of polynomials of the coordinates $x$, $y$ and $z$, the elements of the \textit{rotation vector} $\bvec{r}^\mathsf{T}$ can be computed analytically and only once for a given $l_{max}$ (see \citealt[Eq. 20]{starry}).\\\\

@@

\subsection{Limb darkened surfaces}
A simpler situation is encountered when the surface of the spherical body can be described by a radially-symmetric intensity profile $I(r)$, such as a limb darkened star. In this case, the vector of spherical harmonics coefficients $\bvec{y}$ is very sparse (because all $m\neq0$ components are zero) and the surface map doesn't need to be rotated.\\\\
Contemporary to \cite{starry}, \cite{Agol2020} explore this case and provide simple expressions for the integrated intensity of a star whose surface can be described by a polynomial limb darkening law and occulted by an opaque body. In this case:
\begin{enumerate}
    \item The surface is simply described by $\bvec{u}$, a vector of polynomial limb darkening coefficients.
    \item The Green's basis takes a simpler form (see \citealt[Equation 14]{Agol2020}).
    \item Matrices involved in the computation of the integrated flux have $l_{max}$ columns, versus $(l_{max}+1)^2$ in the general case, reducing drastically the computational cost of the model.
\end{enumerate}
For these reasons, and because of its wide application, our implementation treats radially-symmetric maps separately, leading to highly optimize computations of limb darkened light curves.\\\\

@@

\subsection{Limb darkened maps}
Limb darkened maps are obtained by multiplying a base map, devoid of limb darkening, with a pure limb darkened map. If $\bvec{U}$ is the change of basis matrix that transforms the polynomial limb darkening coefficients $\bvec{u}$ to the spherical harmonics coefficients $\bvec{y}_u$, the pure limb darkened map is
\begin{equation}
    \label{eq:limb_darkening_map}
    \bvec{y}_u = \bvec{U}\bvec{u}.
\end{equation}
The multiplication of the base map represented by $\bvec{y}$ with the limb darkening map represented by $\bvec{y}_u$ can be seen as the multiplication of two polynomials, only with special coefficients (called Clebsch–Gordan coefficients) inherent to the properties of spherical harmonics. To simplify this operation, we first transform both maps into the polynomial basis before multiplying them, so that their product is the actual product of two polynomials which doesn't require the computation of special coefficients.

\subsection{Limb darkening normalization}
In \textsf{starry} version 0.2.2 \citep{starry}, the normalization of a limb-darkened light curve was done by assuming that adding limb darkening to a non-uniform surface does not modify its total luminosity. Hence, the total flux received from a star during its complete rotation should be the same with or without limb darkening. In older versions of starry ($\le$0.2.2), this was done by actually computing the light curve of the star with and without limb darkening, and using the ratio of the two to normalize the limb darkened light curve. As discussed in the \textsf{starry} documentation\footlink{https://starry.readthedocs.io/en/latest/notebooks/LDNormalization/} this approach is wrong and leads to an unphysical normalization.\\\\
Instead, one can assume that the total flux of the star should be similar for all observers, independently of the surface map of the star. In addition to the fact that all components of the spherical harmonics basis integrate to 0 (except the constant component $Y_{0,0}$) we deduce that the mean flux of the star must be coming from a static surface seen by any observer, no matter his viewpoint, which corresponds to a purely limb darkened disk (with no other source of non-uniformity).\\\\
The flux of this purely limb darkened disk is
\begin{equation}
    \label{eq:fu}
    f_u = \bvec{r}^\mathsf{T}\bvec{A_1}\bvec{y}_u
\end{equation}
with $\bvec{y}_u$ defined in \autoref{eq:limb_darkening_map}. Hence, we normalize limb-darkened maps by $\pi/f_u$, so that the total flux for a uniform map with no limb darkening is unchanged.

@@

The rotation of spherical harmonics involves the computation of Wigner-D-matrices, commonly obtained through robust recursion relations in both degree $l$ and order $m$, but leading to costly computations. Hence, knowing how to decompose the full spherical harmonics rotation using pre-computed rotations is essential to achieve optimal performance in this framework. \autoref{fig:rotation_starry} shows the 6 elementary rotations currently used in \textsf{starry}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/figures/rotation_starry.pdf}
        \caption{Decomposition of the full rotation of the surface into elementary rotations. Each schematic shows a sphere rotated from a previous orientation around an axis displayed as a black vector and an angle shown on top. \codelink{workflows/scripts/rotation_figures.py}}
        \label{fig:rotation_starry}
    \end{center}
\end{figure}
In this figure, the first 3 rotations have the effect to rotate the star around its rotation axis, and rotations 3 to 5 place the star on a sky-projected frame. Finally, rotation 6 aligns the spherical harmonics map to the occulting body on the $y$ axis in order to apply Green's theorem in the appropriate basis (Figure 2 from \citealt{starry}).\\\\
The first thing to notice is that rotations 1 to 3 could be simply reduced to a rotation of angle $\theta$ about the y-axis, turning steps 1 to 3 into one. However, rotations around the pole, i.e.\,with a rotation axis along $z$, have simpler expressions that can be implemented separately. In addition, pre-computing these matrices at lower cost is particularly important for steps 2 and 6, involving a potentially large set of angles $\{\theta\}$ and $\{\theta_z\}$, defined over times $\{t\}$, in order to compute the full light curve. This explains the current decomposition of the complete rotation in six separate steps.\\\\
In \textsf{jaxoplanet}, we compute the Wigner-D-matrices by employing the Risbo recursion relations \citep{Risbo1996} implemented in \textsf{JAX} as part of the \textsf{s2fft} Python package \citep{price:s2fft}. In addition, we merge rotations 3, 4 and 5 (see \autoref{fig:rotation_jaxoplanet}) into a single compound rotation of axis 
\begin{equation}
    \label{eq:v}
    \bvec{v} = \frac{1}{\sqrt{1 - \cos^2{\left(\frac{inc}{2} \right)} \cos^2{\left(\frac{obl}{2}\right)}}} \begin{pmatrix}
        \sin{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)}\\
        \sin{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
         - \cos{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
    \end{pmatrix},
\end{equation}
and angle
\begin{equation}
    \label{eq:combined_angle}
    \omega = 2 \cos^{-1}{\left(\cos{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)} \right)}.
\end{equation}
This way, the complete rotation reduces to the four separate steps shown in \autoref{fig:rotation_jaxoplanet}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/figures/rotation_jaxoplanet_1.pdf}
        \caption{Consecutive rotations of the spherical harmonics basis in the \textsf{jaxoplanet} implementation. Each schematic shows a sphere rotated from a previous orientation around an axis displayed as a black vector and an angle shown on top. \codelink{workflows/scripts/rotation_figures.py}}
        \label{fig:rotation_jaxoplanet}
    \end{center}
\end{figure}

@@

\subsubsection{Polynomial limb darkening}
For a surface described by a polynomial limb darkening law of order $n$, the $\mathcal{P}$ integral is given in \cite{Agol2020} by
\begingroup\makeatletter\def\f@size{9}\check@mathfonts
$$
    \mbox{\normalsize$\mathcal{P}(\bvec{G}_n)$} =
    \begin{dcases}
        %
        \pi - \kappa_1 - r^2\kappa_0 + \textit{A}_{kite} & \qquad n = 0
        \\[1em]
        \text{see below} & \qquad n = 1
        \\[1em]
        2s_0 + 4\pi\eta - 2\pi & \qquad n = 2
        \\[1em]
        2r(4br)^{\frac{n}{2}}
        \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        (k^2 - \sin^2{\varphi '})^{\frac{n}{2}}
        (r - b + 2b\sin^2\varphi ') \, \text{d}\varphi ' & \qquad \mathrm{n\ge3},
    \end{dcases}
$$
\endgroup
where $\kappa_0$, $\kappa_1$ and $A_{kite}$ are the angles and area shown in \autoref{fig:geometry} and computed following \cite{Agol2020} (Equations 31 and 32). For $n=2$, $\eta$ is defined in \cite{Agol2020} (Equation 53) and depends on the elliptic parameter
%
\begin{equation}
    \label{eq:k}
    k = \frac{1-r^2 - b^2 + 2br}{4br},
\end{equation}\\
%
also used in the case $n=3$. For $n=1$, we use Equation 31, 32 and 34 from \cite{starry} to write
%
\begin{equation}
    \mathcal{P}(\bvec{G}_1) = \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        r(r-b\cos{2\varphi '})\;\frac{1-z^3}{3(1-z^2)}\text{d}\varphi ' \quad \text{and} \quad \mathcal{Q}(\bvec{G}_1) = \frac{2}{3}(\pi-\kappa_1),
\end{equation}\\
%
where $1 - z^2 = b + r - 2\cos{\varphi '}$ \footnote{note that $\varphi '$ is different from the angle $\varphi$ shown in \autoref{fig:geometry} and was introduced through the change of variable $\varphi ' = \frac{1}{2}(\varphi -\frac{3\pi}{2})$.}.
\subsubsection{Non-uniform surface}\label{non-uniform}
In the general case, for a star with a non-radially symmetric surface intensity, $n$ denotes the index of the spherical harmonics component such that
\begin{equation}
    \label{eq:n}
    n = l^2 + l + m,
\end{equation}
where $l$ is the degree of the spherical harmonic and $m$ its order. For each component, the expression of $\mathcal{P}(\bvec{G}_n)$ depends on $\mu$ and $\nu$ defined as
\begin{equation}
    \label{eq:munu}
    \begin{aligned}
    \mu &\equiv l - m,\\
    \nu &\equiv l + m.
    \end{aligned}
\end{equation}

The $\mathcal{P}$ integral is then given by Equation D32 from \cite{starry} as
%
\begingroup\makeatletter\def\f@size{9}\check@mathfonts
$$
    \mbox{\normalsize$\mathcal{P}(\bvec{G}_n)$} =
    \begin{dcases}
        %
        2(2r)^{l+2}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2 -s_{\varphi '}^4)^{\frac{\mu+4}{4}}
            (\delta + s_{\varphi '}^2)^{\frac{\nu}{2}}
            \, \text{d}\varphi '
            %
            & \qquad \frac{\mu}{2} \, \mathrm{even}
        \\[1em]
        %
        \mathcal{F}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2-s_{\varphi '}^4)^{\tfrac{l-2}{2}}
            (k^2- s_{\varphi '}^2)^{\frac{3}{2}}
            (1-2s_{\varphi '}^2)
            \, \text{d}\varphi '
            %
            & \qquad \mu = 1, \,
                     l \, \mathrm{even}
        \\[1em]
        %
        \mathcal{F}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2-s_{\varphi '}^4)^{\tfrac{l-3}{2}}
            (\delta + s_{\varphi '}^2)
            (k^2- s_{\varphi '}^2)^{\frac{3}{2}}
            (1-2 s_{\varphi '}^2)
            \, \text{d}\varphi '
            %
            & \qquad \mu = 1, \, l \ne 1, \,
                     l \, \mathrm{odd}
        \\[1em]
        %
        2\mathcal{F}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2-s_{\varphi '}^4)^{\frac{\mu-1}{4}}
            (\delta + s_{\varphi '}^2)^{\frac{\nu-1}{2}}
            (k^2 - s_{\varphi '}^2)^{\frac{3}{2}}
            \, \text{d}\varphi '
            & \qquad \frac{\mu-1}{2} \, \mathrm{even}, \, l \ne 1
        \\[1em]
        %
        \int\displaylimits_{-\kappa/2}^{\kappa/2}
        r(r-b\cos{2\varphi '})\;\frac{1-z^3}{3(1-z^2)}\text{d}\varphi ' & \qquad \mu=1, \, l=1
        \\[1em]
        %
        0 & \qquad \mathrm{otherwise},
    \end{dcases}
$$
\endgroup
%
where $s_{\varphi '} = \sin{\varphi '}$, $\kappa = \varphi + \frac{\pi}{2}$ (see \citealt[Equation D31]{starry}) and $\delta$ is defined for numerical stability as
\begin{equation}
    \label{eq:delta}
    \delta=\frac{b-r}{2r}.
\end{equation}
Note how, unlike in the \textsf{starry} implementation, the term $\mu=1$, $l=1$ is derived exactly like in the polynomial limb darkening case, instead of analytically which would require the evaluation of elliptic integrals (see \citealt{starry}, Section D.2.3).\\\\
Because of its numerical integration, the precision of the $\mathcal{P}$ integral is limited and depends on the order $q$ of the Gauss-Legendre quadrature, which corresponds to the number of point for which the integrand is evaluated. To minimize this number, we note that all integrands are symmetrical functions, except for $\mu=1$ and $l=1$. Hence, these integrals can be computed over the interval $\left[0, \frac{\kappa}{2}\right]$, instead of $\left[-\frac{\kappa}{2}, \frac{\kappa}{2}\right]$, and multiplied by 2. We also note that all terms for which $\frac{\mu}{2}$ is even reach machine precision for an integration order as low as $q=20$, that we hold fix. With these optimizations, we benchmark the precision and speed of our final implementation for the solution vector and the total flux in \autoref{performance}.

>>>>

% author info
\author[0000-0002-4296-2246]{Lionel J. Garcia}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY, USA}
\author[0000-0002-0842-863X]{Soichiro Hattori}
\affiliation{Department of Astrophysics, American Museum of Natural History, New York, NY, USA}
\affiliation{Department of Astronomy, Columbia University, New York, NY, USA}
\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY, USA}

@@

\begin{abstract}
    Modeling light curves of celestial bodies serves a wide variety of science cases. From the orbital characterization of stellar systems to the study of exoplanet and stellar atmospheres. In this context, the framework known as \textit{starry} provided the community with a powerful tool to efficiently and accurately compute light curves of bodies with non-uniform surface intensities. However, as datasets and models complexity grows, the tractable inference of their physical parameters becomes increasingly challenging. In this paper, we present a new implementation of \textsf{starry} built from scratch in \textsf{JAX}, a high-performance machine-learning library designed for large-scale numerical computation. We describe the changes made to the original version of \textsf{starry} and evaluate the performance of our new implementation that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a fast and differentiable model of stellar light curves, and enable the use of powerful tools available in the \textsf{JAX} ecosystem. \gitlink{}
\end{abstract}

@@

In the age of wide-field surveys (such as the \textit{Transiting Exoplanet Survey Satellite}; \citealt{tess}) and high-precision spectroscopic instruments, these models are being compared to large datasets in probabilistic ways, requiring fast and robust evaluations of their likelihood over larger and larger parameter spaces. These inferences can be facilitated by gradient-based optimization and sampling algorithms (see for example \citealt{Betancourt2017}), so that being able to differentiate these models with respect to their parameters offers clear advantages.\\\\
In this context, \cite{starry} and \cite{Agol2020} presented a framework to analytically compute the light curves of spherical bodies with non-uniform surface intensities, such as stars and exoplanets, and the flux observed from their mutual occultations. This framework, known as \textit{starry}, has been successfully applied to a wide variety of science cases, like the study of transiting exoplanets \citep{Kostov2019}, precise modeling of radial velocities \citep{Dawson2021}, the study of exoplanets atmospheres \citep{Bell2023}, the study of stellar photospheres \citep{Almenara2022}, or even the surface mapping of stellar system objects \citep{Bartolic2022}. However, as datasets and models complexity grows, tractable inference becomes increasingly challenging, and often necessitates biased approximations. This is the case, for example, in transmission spectroscopy measurements conducted with the \textit{James Webb Space Telescope} NIRSpec instrument, that should ideally consist in the simultaneous modeling of more than 3000 transit light curves, one per pixel element (e.g. \citealt{Alderson2023}).\\\\
In this paper we present \textsf{jaxoplanet}, a new implementation of the \textit{starry} framework in \textsf{JAX}, a high-performance library designed for large-scale computations, including autodifferentiation capabilities, with code compiled and deployable on various hardware accelerators. In \autoref{starry}, we summarize the formalism of \textit{starry}, introduced by \cite{starry} to analytically compute stellar light curves. In \autoref{optimization}, we describe the \textsf{jaxoplanet} implementation and the changes made to the \textsf{starry} implementation. Finally, in \autoref{performance}, we evaluate the performance of our new implementation that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a differentiable model of stellar light curves, and enable the use of powerful machine-learning tools available in the \textsf{JAX} ecosystem.\\\\

@@

\cite{starry} and \cite{Agol2020} present a general framework to compute the light curves of spherical bodies with non-uniform surface intensities called \textit{starry}. In this section, we describe the mathematical formalism behind this model.

@@

In order to simplify this integral, \citealt{starry} introduce a change of basis from the spherical harmonics basis $\bvec{\tilde y}$ to a simple polynomial basis $\bvec{\tilde p}$ (defined in \citealt[section 2.3]{starry}) characterized by the change of basis matrix $\bvec{A_1}$. In this new basis 

@@

As $\bvec{\tilde{p}}$ is simply made of polynomials of the coordinates $x$, $y$ and $z$, the elements of the \textit{rotation vector} $\bvec{r}^\mathsf{T}$ can be computed in closed form and only once for a given $l_{max}$ (see \citealt[Eq. 20]{starry}).\\\\

@@

\subsection{Limb darkened surfaces}
A simpler situation is encountered when the surface of the spherical body can be described by a radially-symmetric intensity profile $I(r)$, such as a limb darkened star. In this case, the vector of spherical harmonics coefficients $\bvec{y}$ is very sparse (because all $m\neq0$ components are zero) and the surface map doesn't need to be rotated.\\\\
Contemporary to \cite{starry}, \cite{Agol2020} explore this case and provide simple expressions for the integrated intensity of a star whose surface can be described by a polynomial limb darkening law and occulted by an opaque body. In this case:
\begin{enumerate}
    \item The surface is simply described by $\bvec{u}$, a vector of polynomial limb darkening coefficients.
    \item The Green's basis takes a simpler form (see \citealt[Equation 14]{Agol2020}).
    \item Matrices involved in the computation of the integrated flux have $l_{max}$ columns, versus $(l_{max}+1)^2$ in the general case, drastically reducing the computational cost of the model.
\end{enumerate}
For these reasons, and because of its wide application, our implementation treats radially-symmetric maps separately, leading to highly optimized computations of limb darkened light curves.\\\\

@@

\subsection{Limb darkened maps}
Limb darkened maps are obtained by multiplying a base map, devoid of limb darkening, with a pure limb darkened map. If $\bvec{\tilde u}$ is the change of basis matrix that transforms the polynomial limb darkening coefficients $\bvec{u}$ to the spherical harmonics coefficients $\bvec{y}_u$, the pure limb darkened map is
\begin{equation}
    \label{eq:limb_darkening_map}
    \bvec{y}_u = \bvec{\tilde u}^\mathsf{T} \bvec{u}.
\end{equation}
The multiplication of the base map represented by $\bvec{y}$ with the limb darkening map represented by $\bvec{y}_u$ can be seen as the multiplication of two polynomials, only with special coefficients called Clebsch–Gordan coefficients. To simplify this operation, we first transform both maps into the polynomial basis before multiplying them, so that their product is the actual product of two polynomials which doesn't require the computation of special coefficients.

\subsection{Limb darkening normalization}
In \textsf{starry} version 0.2.2 \citep{starry}, the normalization of a limb-darkened light curve was done by assuming that adding limb darkening to a non-uniform surface does not modify its total luminosity. Hence, the total flux received from a star during its complete rotation should be the same with or without limb darkening. In older versions of starry ($\le$0.2.2), this was done by actually computing the light curve of the star with and without limb darkening, and using the ratio of the two to normalize the limb darkened light curve. As discussed in the \textsf{starry} documentation\footlink{https://starry.readthedocs.io/en/latest/notebooks/LDNormalization/} this approach is wrong and leads to an unphysical normalization.\\\\
As we notice that all components of the spherical harmonics basis integrate to 0 (except the constant component $Y_{0,0}$) we deduce that the mean flux of the star must be coming from a static surface seen by any observer, no matter his viewpoint, which corresponds to a purely limb darkened disk (with no other source of non-uniformity).\\\\
The flux of this purely limb darkened disk is
\begin{equation}
    \label{eq:fu}
    F_u = \bvec{r}^\mathsf{T}\bvec{A_1}\bvec{y}_u
\end{equation}
with $\bvec{y}_u$ defined in \autoref{eq:limb_darkening_map}. Hence, we normalize limb-darkened maps by $\pi/F_u$, so that the total flux for a uniform map with no limb darkening is unchanged.

@@

The rotation of spherical harmonics involves the computation of Wigner-D-matrices, commonly obtained through robust recursion relations in both degree $l$ and order $m$, but leading to costly computations. Hence, knowing how to decompose the full spherical harmonics rotation using pre-computed rotations is essential to achieve optimal performance in this framework. \autoref{fig:rotation_starry} shows the 6 elementary rotations currently used in \textsf{starry}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/figures/rotation_starry.pdf}
        \caption{Decomposition of the full rotation of the surface into elementary rotations. Each schematic illustrates a sphere undergoing rotation by the angle specified above it. Arrows indicate the approximate movement of the sphere between successive configurations. The numbers at the bottom of each sphere represent the index of each rotation. \codelink{workflows/scripts/rotation_figures.py}}
        \label{fig:rotation_starry}
    \end{center}
\end{figure}
In this figure, the first 3 rotations have the effect to rotate the star around its rotation axis, and rotations 3 to 5 place the star on a sky-projected frame. Finally, rotation 6 aligns the spherical harmonics map to the occulting body on the $y$ axis in order to apply Green's theorem in the appropriate basis (Figure 2 from \citealt{starry}).\\\\
The first thing to notice is that rotations 1 to 3 could be simply reduced to a rotation of angle $\theta$ about the y-axis, turning steps 1 to 3 into one. However, rotations around the pole, i.e.\,with a rotation axis along $z$, have simpler expressions that can be implemented separately. In addition, pre-computing these matrices at lower cost is particularly important for steps 2 and 6, involving a potentially large set of angles $\{\theta\}$ and $\{\theta_z\}$, defined over times $\{t\}$, in order to compute the full light curve. This explains the current decomposition of the complete rotation in six separate steps.\\\\
In \textsf{jaxoplanet}, we compute the Wigner-D-matrices by employing the Risbo recursion relations \citep{Risbo1996} implemented in \textsf{JAX} as part of the \textsf{s2fft} Python package \citep{price:s2fft}. In addition, we merge rotations 3, 4 and 5 (see \autoref{fig:rotation_jaxoplanet}) into a single compound rotation of axis 
\begin{equation}
    \label{eq:v}
    \bvec{v} = \frac{1}{\sqrt{1 - \cos^2{\left(\frac{inc}{2} \right)} \cos^2{\left(\frac{obl}{2}\right)}}} \begin{pmatrix}
        \sin{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)}\\
        \sin{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
         - \cos{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
    \end{pmatrix},
\end{equation}
and angle
\begin{equation}
    \label{eq:combined_angle}
    \omega = 2 \cos^{-1}{\left(\cos{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)} \right)}.
\end{equation}
This way, the complete rotation reduces to the four separate steps shown in \autoref{fig:rotation_jaxoplanet}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/figures/rotation_jaxoplanet_1.pdf}
        \caption{Consecutive rotations of the spherical harmonics basis in the \textsf{jaxoplanet} implementation. Each schematic illustrates a sphere undergoing rotation by the angle specified above it. Arrows indicate the approximate movement of the sphere between successive configurations. The numbers at the bottom of each sphere represent the index of each rotation. \codelink{workflows/scripts/rotation_figures.py}}
        \label{fig:rotation_jaxoplanet}
    \end{center}
\end{figure}

@@

\subsubsection{Polynomial limb darkening}
For a surface described by a polynomial limb darkening law of order $n$, the $\mathcal{P}$ integral is given in \cite{Agol2020} by
\begingroup\makeatletter\def\f@size{9}\check@mathfonts
$$
    \mbox{\normalsize$\mathcal{P}(\bvec{G}_n)$} =
    \begin{dcases}
        %
        \pi - \kappa_1 - r^2\kappa_0 + \textit{A}_{kite} & \qquad n = 0
        \\[1em]
        \text{see below} & \qquad n = 1
        \\[1em]
        2\,\mathcal{P}(\bvec{G}_0) + 4\pi\eta - 2\pi & \qquad n = 2
        \\[1em]
        2r(4br)^{\frac{n}{2}}
        \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        (k^2 - \sin^2{\varphi '})^{\frac{n}{2}}
        (r - b + 2b\sin^2\varphi ') \, \text{d}\varphi ' & \qquad \mathrm{n\ge3},
    \end{dcases}
$$
\endgroup
where $\kappa_0$, $\kappa_1$ and $A_{kite}$ are the angles and area shown in \autoref{fig:geometry} and computed following \cite{Agol2020} (Equations 31 and 32). For $n=2$, $\eta$ is defined in \cite{Agol2020} (Equation 53) and depends on the elliptic parameter
%
\begin{equation}
    \label{eq:k}
    k = \frac{1-r^2 - b^2 + 2br}{4br},
\end{equation}\\
%
also used in the case $n=3$. For $n=1$, we use Equation 31, 32 and 34 from \cite{starry} to write
%
\begin{equation}
    \mathcal{P}(\bvec{G}_1) = \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        r(r-b\cos{2\varphi '})\;\frac{1-z^3}{3(1-z^2)}\text{d}\varphi ' \quad \text{and} \quad \mathcal{Q}(\bvec{G}_1) = \frac{2}{3}(\pi-\kappa_1),
\end{equation}\\
%
where $1 - z^2 = b + r - 2\cos{\varphi '}$ \footnote{note that $\varphi '$ is different from the angle $\varphi$ shown in \autoref{fig:geometry} and was introduced through the change of variable $\varphi ' = \frac{1}{2}(\varphi -\frac{3\pi}{2})$.}.
\subsubsection{Non-uniform surface}\label{non-uniform}
In the general case, for a star with a non-radially symmetric surface intensity, $n$ denotes the index of the spherical harmonics component such that
\begin{equation}
    \label{eq:n}
    n = l^2 + l + m,
\end{equation}
where $l$ is the degree of the spherical harmonic and $m$ its order. For each component, the expression of $\mathcal{P}(\bvec{G}_n)$ depends on $\mu$ and $\nu$ defined as
\begin{equation}
    \label{eq:munu}
    \begin{aligned}
    \mu &\equiv l - m,\\
    \nu &\equiv l + m.
    \end{aligned}
\end{equation}

The $\mathcal{P}$ integral is then given by Equation D32 from \cite{starry} as
%
\begingroup\makeatletter\def\f@size{9}\check@mathfonts
$$
    \mbox{\normalsize$\mathcal{P}(\bvec{G}_n)$} =
    \begin{dcases}
        %
        2(2r)^{l+2}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2 -s_{\varphi '}^4)^{\frac{\mu+4}{4}}
            (\delta + s_{\varphi '}^2)^{\frac{\nu}{2}}
            \, \text{d}\varphi '
            %
            & \qquad \frac{\mu}{2} \, \mathrm{even}
        \\[1em]
        %
        \mathcal{F}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2-s_{\varphi '}^4)^{\tfrac{l-2}{2}}
            (k^2- s_{\varphi '}^2)^{\frac{3}{2}}
            (1-2s_{\varphi '}^2)
            \, \text{d}\varphi '
            %
            & \qquad \mu = 1, \,
                     l \, \mathrm{even}
        \\[1em]
        %
        \mathcal{F}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2-s_{\varphi '}^4)^{\tfrac{l-3}{2}}
            (\delta + s_{\varphi '}^2)
            (k^2- s_{\varphi '}^2)^{\frac{3}{2}}
            (1-2 s_{\varphi '}^2)
            \, \text{d}\varphi '
            %
            & \qquad \mu = 1, \, l \ne 1, \,
                     l \, \mathrm{odd}
        \\[1em]
        %
        2\mathcal{F}\int\displaylimits_{-\kappa/2}^{\kappa/2}
            (s_{\varphi '}^2-s_{\varphi '}^4)^{\frac{\mu-1}{4}}
            (\delta + s_{\varphi '}^2)^{\frac{\nu-1}{2}}
            (k^2 - s_{\varphi '}^2)^{\frac{3}{2}}
            \, \text{d}\varphi '
            & \qquad \frac{\mu-1}{2} \, \mathrm{even}, \, l \ne 1
        \\[1em]
        %
        \int\displaylimits_{-\kappa/2}^{\kappa/2}
        r(r-b\cos{2\varphi '})\;\frac{1-z^3}{3(1-z^2)}\text{d}\varphi ' & \qquad \mu=1, \, l=1
        \\[1em]
        %
        0 & \qquad \mathrm{otherwise},
    \end{dcases}
$$
\endgroup
%
where $s_{\varphi '} = \sin{\varphi '}$, $\kappa = \varphi + \frac{\pi}{2}$ (see \citealt[Equation D31]{starry}) and $\delta$ is defined for numerical stability as
\begin{equation}
    \label{eq:delta}
    \delta=\frac{b-r}{2r}.
\end{equation}
Note how, unlike in the \textsf{starry} implementation, the term $\mu=1$, $l=1$ is derived exactly like in the polynomial limb darkening case, instead of analytically which would require the evaluation of elliptic integrals (see \citealt{starry}, Section D.2.3).\\\\
Because of its numerical integration, the precision of the $\mathcal{P}$ integral is limited and depends on the order $q$ of the Gauss-Legendre quadrature, which corresponds to the number of point for which the integrand is evaluated. To minimize this number, we note that all integrands are symmetrical functions, except for $\mu=1$ and $l=1$. Hence, these integrals can be computed over the interval $\left[0, \frac{\kappa}{2}\right]$, instead of $\left[-\frac{\kappa}{2}, \frac{\kappa}{2}\right]$, and multiplied by 2. We also note that all terms for which $\frac{\mu}{2}$ is even reach machine precision for an integration order as low as $q=20$, that we hold fixed. With these optimizations, we benchmark the precision and speed of our final implementation for the solution vector and the total flux in \autoref{performance}.
